miniKanren with fair search strategies

### 2019-05-06 Paper/paper-slatex.tex
fair conj

miniKanren is a family of relational programming languages. miniKanren
programs, especially relational interpreters, have been proven to be
useful in solving many problems by \citet{byrd2017unified}. A subtlety in 
writing miniKanren programs with large \scheme|conde| expressions, such as 
relational interpreters, is that the order of \scheme|conde| clauses sometimes 
affect the speed considerably. This phenomenon also appears when running 
miniKanren programs with the implementation in \citet{friedman_reasoned_2018}, 
one of the most well-understood implementation. This is because the search 
strategy of this implementation, interleaving depth-first search (\dfsi), gives 
left clauses higher ``search priority'' when a \scheme|conde| has more than two 
clauses. The situation is exacerbated as the number of clauses increases. This 
unfair treatment causes two problems when a \scheme|conde| expression has many 
clauses: the right-most clause can hardly contribute to query result; and 
programmers might need to tweak the clause order to maximize their programs' 
efficiency, depending on the distribution of queries.

@@

Before going further into fairness, we would like to give a short review about 
state, search space, and goal, because fairness is defined in terms of them. 
Every answer given by miniKanren corresponds to a state. Each state records a 
way how some relations can be satisfied. In our case, the state is just a 
subsitution associating logic variables with terms. In more expressive 
miniKanrens (e.g. miniKanren with with symbolic constraints 
\citep{hemann2017framework}), states can include more information. A search 
space is a collection of states. And a goal is a function from a state to a 
search space. Goals check their inputs and produces possibly extended states. A 
goal might fail for some input, in which case the output search space would be 
empty.

>>>>

miniKanren is a family of relational programming languages. miniKanren
programs, especially relational interpreters, have been proven to be
useful in solving many problems by \citet{byrd2017unified}.

A subtlety in writing miniKanren programs with large \scheme|conde| expressions, 
such as relational interpreters, is that the order of \scheme|conde| clauses 
sometimes affect the speed considerably. This phenomenon also appears when 
running miniKanren programs with the implementation in 
\citet{friedman_reasoned_2018}, one of the most well-understood implementation. 
This is because the search strategy of this implementation, interleaving 
depth-first search (\dfsi), gives left clauses higher ``search priority'' when a 
\scheme|conde| has more than two clauses. The situation is exacerbated as the 
number of clauses increases. This unfair treatment causes two problems when a 
\scheme|conde| expression has many clauses: the right-most clause can hardly 
contribute to query result; and programmers might need to tweak the clause order 
to maximize their programs' efficiency, depending on the distribution of 
queries.

@@

Before going further into fairness, we would like to give a short review about 
state, search space, and goal, because fairness is defined in terms of them. 
Every answer given by miniKanren corresponds to a state. Each state records a 
way how some relations can be satisfied. In our case, the state is just a 
subsitution associating logic variables with terms. In more expressive 
miniKanrens (e.g. miniKanren with with symbolic constraints 
\citep{hemann2017framework}), states can include more information. A search 
space is a collection of states. And a goal is a function from a state to a 
search space. Goals check their input states and outputs possibly extended 
states. A goal might fail for some inputs, in which case the output search 
space would be empty.

### 2019-05-06 Paper/paper.tex
fair conj

\newcommand{\conde}{\texttt{cond$^e$}}
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}
\newcommand{\DFSi}[0]{DFS$_\textrm{i}$}
\newcommand{\DFSf}[0]{DFS$_\textrm{f}$}
\newcommand{\BFS }[0]{BFS}

@@

The miniKanren in TR2 implements interleaving DFS(\DFSi), the cause of this 
unexpected result. With this search strategy, each \conde clause takes a half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
recourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. DFS$_\textrm{i}$, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde contains more clauses.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-iDFS.rkt}
	\end{tabular}
\end{center}

Under the hood, the \conde here is allocating computational effort to 
four trivially different search space. The unfair \disj{} in 
\DFSi{} allocates much more effort to the first search space. On the contrary, 
fair \disj{} would allocate effort evenly to each search space.

@@

Running the same program again with almost-fair \disj 
(e.g. DFS$_\textrm{bi}$) gives the same result. Almost-fair, however, is not 
completely fair, as shown by the following example.

@@

DFS$_\textrm{bi}$ is fair only when the number of goals is a power of 2, 
otherwise, some goals would be allocated twice as many resources than the 
others. In the above example, where the \conde has five clauses, the clauses of 
\texttt{b}, \texttt{c}, and \texttt{d} are allocated more resources.

We end this subsection with precise definitions of all levels of 
\disj{} fairness. Our definition of \emph{fair} \disj{} is slightly generalize 
from the one by \citet{seres1999algebra}. Their definition is for binary 
disjunct. We generalize it to multi-arity one.

\begin{defn}[fair \disj{}]
	A \disj{} is fair iff it allocates computational resource evenly to 
	search spaces produced by goals in the same disjunct (e.g. clauses in the 
	same 
	\conde).
\end{defn}

\begin{defn}[almost-fair \disj{}]
	A \disj{} is almost-fair iff it allocates computational resource so 
	evenly to search spaces produced by goals in the same disjunct that the 
	maximal 
	ratio of resources is bounded by a constant.
\end{defn}

@@

In the following program, the three \conde clauses differ in a trivial way. So 
we expect lists of each letter constitute $1/4$ of the answer list. Search 
strategies with unfair \conj{} (e.g. DFS$_\textrm{i}$, DFS$_\textrm{f}$), 
however, give us many more lists of \texttt{a}s than lists of other letters. 
And some letters (e.g. lists 
of \texttt{c}) are rarely found. Although DFS$_\textrm{i}$'s \disj{} is unfair in general, 
it is fair when there is no call to relational definition in sub-goals, 
including this case. The situation would be worse if we add more \conde 
clauses. The result with DFS$_\textrm{bi}$, whose \conj{} is also unfair, is similar, but 
due to its different \disj{}, the position of \texttt{b} and \texttt{c} are 
swapped.

@@

Intuitively, search strategies with fair \conj{} should produce each letter of 
lists equally frequently. Indeed, BFS does so.

@@

A more interesting situation is when the first conjunctive goal produces 
infinite many answers. Consider the following example, a naive specification of 
fair \conj{} might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}.

\NOTE[MVC]{incomplete w.r.t. what? where's the definition of incomplete?}

@@

Our solution requires a search strategy with \emph{fair} \conj{} to 
package answers in bags, where each bag contains finite answers, and to allocate 
resources evenly among search spaces derived from answers in the same bag. The 
way to package depends on search strategy. And how to allocate resources among 
search space related to different bags is unspecified. Our definition of fair 
\conj{} is orthogonal with completeness. For example, a naively fair strategy 
is fair but not complete, while BFS is fair and complete. 
\NOTE[MVC]{Also here, complete w.r.t what?}

BFS packages answers by their costs. The \emph{cost} of a answer is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citet{seres1999algebra}. In the following example, 
every answer is a list of list of symbol. The cost of each of them is equal to 
the length of the inner lists plus the length of the outer list. In addition to 
being fair, BFS also produces answers in increasing order of cost.

@@

%TODO describe DFSi here

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair \conj{}. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}s, hence 
the name of this search strategy. In contrast, the \disj{} in DFS$_\textrm{i}$
constructs the binary tree with the same nodes but in the unbalanced form.

@@

Fair DFS (DFS$_\textrm{f}$) has fair \disj{} and unfair \conj{}. The implementation of DFS$_\textrm{f}$ 
differs from DFS$_\textrm{i}$'s in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is 
changed to call a new and fair version of \texttt{append-inf}. 
\texttt{append-inf/fair} immediately calls its helper, 
\texttt{append-inf/fair\^{}}, with the first argument, \texttt{s?}, set to 
\texttt{\#{}t}, which indicates that \texttt{s-inf} and \texttt{t-inf} haven't 
been swapped. The swapping happens at the third \texttt{cond} clause in the 
helper, where \texttt{s?} is updated accordingly. The first two \texttt{cond} 
clauses essentially copy the \texttt{car}s and stop recursion when one of the 
input spaces is obviously finite. The third clause, as we mentioned above, is 
just for swapping. When the fourth and last clause runs, we know that both 
\texttt{s-inf} and \texttt{t-inf} are ended with a thunk. In this case, a new 
thunk is constructed. The new thunk calls the driver recursively. Here changing 
the order of \texttt{t-inf} and \texttt{s-inf} won't hurt the fairness (though 
it will change the order of answers). We swapped them back so that answers are 
produced in a more natural order.

@@

\NOTE[MVC]{Though bag is well known, people rarely say ``bagging''. How about putting information in a bag, or something better?}
\NOTE[MVC]{What is the bagging information?}
\NOTE[LKC]{It's just cost... You're right. I should have be more direct.}

>>>>

\newcommand{\conde}{\texttt{cond$^e$}}
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}
\newcommand{\DFSi }[0]{DFS$_\textrm{i}$}
\newcommand{\DFSf }[0]{DFS$_\textrm{f}$}
\newcommand{\DFSbi}[0]{DFS$_\textrm{bi}$}
\newcommand{\BFS  }[0]{BFS}

@@

The miniKanren in TR2 implements interleaving DFS(\DFSi), the cause of this 
unexpected result. With this search strategy, each clause takes a half 
of its received computational resources and pass the other half to its 
following clauses, except for the last clause that takes all resources it 
receives. In the example above, the \texttt{a} clause takes half of all 
recourses. And the \texttt{b} clause takes a quarter. Thus \texttt{c} and 
\texttt{d} barely contribute to the result.

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. \DFSi, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters  (e.g. \texttt{c} and \texttt{d}) are
rarely seen. The situation would be exacerbated if \conde contains more clauses.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-disj-DFSi.rkt}
	\end{tabular}
\end{center}

Under the hood, the \conde here is allocating computational resource to 
four trivially different search space. The unfair \disj{} in 
\DFSi{} allocates many more resources to the first search space. On the 
contrary, fair \disj{} would allocate resources evenly to each search space.

@@

Running the same program again with almost-fair \disj {} (e.g. 
DFS$_\textrm{bi}$) gives the same result. Almost-fair, however, is not 
completely fair, as shown by the following example.

@@

\DFSbi{} is fair only when the number of goals is a power of 2, 
otherwise, some goals would be allocated twice as many resources than the 
others. In the above example, where the \conde has five clauses, the clauses of 
\texttt{b}, \texttt{c}, and \texttt{d} are allocated more resources.

We end this subsection with precise definitions of all levels of 
\disj{} fairness. Our definition of \emph{fair} \disj{} is slightly generalize 
from the one by \citet{seres1999algebra}. Their definition is for binary 
disjunction. We generalize it to multi-arity one.

\begin{defn}[fair \disj{}]
A \disj{} is fair iff it allocates computational resource evenly to 
search spaces produced by goals in the same disjunction (i.e. clauses in 
the same \conde).
\end{defn}

\begin{defn}[almost-fair \disj{}]
A \disj{} is almost-fair iff it allocates computational resource so evenly to 
search spaces produced by goals in the same disjunction that the maximal ratio 
of resources is bounded by a constant.
\end{defn}

@@

Given the following program, it is natural to expect lists of each letter to
constitute $1/4$ in the answer. Search strategies with unfair \conj{} (e.g. 
\DFSi, \DFSbi, \DFSf), however, results in many more lists of \texttt{a}s than 
lists of other letters. And some letters are rarely seen. The situation would 
be exacerbated if \conde contains more clauses.

\NOTE[LKC]{Should I add a footnote?

``
Although 
DFS$_\textrm{i}$'s \disj{} is unfair in general, it is fair when there is no 
call to relational definition in \conde{} clauses (including this case).
''}

@@

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-DFSf.rkt}
	\end{tabular}
\end{center}

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-DFSbi.rkt}
	\end{tabular}
\end{center}

Under the hood, the \conde{} and the call to \texttt{repeato} are connected by 
\conj{}. The \conde{} goal outputs a search space including four trivially 
different states. These states are then supplied individually to 
\texttt{(repeato x q)} goal, producing four search spaces. In the examples 
above, the \conj{}s are allocating more computational resources to the state of 
\texttt{a}. On the contrary, fair \conj{} would allocate resources evenly to 
each search space.

@@

A more interesting situation is when the first conjunct produces infinite many 
answers. Consider the following example, a naive specification of fair \conj{} 
might require search strategies to produce all sorts of singleton lists, but no 
longer ones, which makes the strategies incomplete. A search strategy is 
\emph{complete} iff it can find out all the answers in finite time, otherwise, 
it is \emph{imcomplete}.

@@

Our solution requires a search strategy with \emph{fair} \conj{} to organize
states in bags in search spaces, where each bag contains finite states, and 
to allocate resources evenly among search spaces derived from states in the 
same bag. It is up to a search strategy designer to decide by what criteria to 
put states in the same bag, and how to allocate resources among search spaces 
related to different bags.

BFS puts states of the same cost in the same bag, and allocate resources 
carefully among search spaces related to different bags such that answers are 
produced in increasing order of cost. The \emph{cost} of a answer is its depth 
in the search tree (i.e. the number of calls to relational definitions required 
to find them) \citet{seres1999algebra}. In the following example, the cost of 
each answer is equal to the length of the inner lists plus the length of the 
outer list.

@@

We end this subsection with precise definitions of all levels of \conj{} 
fairness.

% Our definition of fair \conj{} is orthogonal with completeness. For 
% example, a naively fair strategy is fair but not complete, while BFS is fair and 
% complete.

\begin{defn}[fair \conj{}]
A \conj{} is fair iff it allocates computational resource evenly to 
search spaces produced from states in the same bag. A bag is a finite 
collection of state. And search strategies with fair \conj{} should represent 
search spaces with possibly infinite collection of state. 
\end{defn}

\begin{defn}[unfair \conj{}]
A \conj{} is unfair iff it is not fair.
\end{defn}

\section{interleaving depth-first search}

In this section we review the implementation of interleaving depth-first search 
(\DFSi). ... ...

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair \conj{}. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}s, hence 
the name of this search strategy. In contrast, the \disj{} in DFS$_\textrm{i}$
constructs the binary tree with the same nodes but in an unbalanced form.

@@

Fair DFS (DFS$_\textrm{f}$) has fair \disj{} and unfair \conj{}. The implementation of DFS$_\textrm{f}$ 
differs from DFS$_\textrm{i}$'s in \texttt{disj2} (Fig.~\ref{fDFS}). \texttt{disj2} is 
changed to call a new and fair version of \texttt{append-inf}. 
\texttt{append-inf/fair} immediately calls its helper, \texttt{loop}, with the 
first argument, \texttt{s?}, set to \texttt{\#{}t}, which indicates that 
\texttt{s-inf} and \texttt{t-inf} haven't been swapped. The swapping happens at 
the third \texttt{cond} clause in the helper, where \texttt{s?} is updated 
accordingly. The first two \texttt{cond} clauses essentially copy the 
\texttt{car}s and stop recursion when one of the input spaces is obviously 
finite. The third clause, as we mentioned above, is just for swapping. When the 
fourth and last clause runs, we know that both \texttt{s-inf} and \texttt{t-inf} 
are ended with a thunk. In this case, a new thunk is constructed. The new thunk 
calls the driver recursively. Here changing the order of \texttt{t-inf} and 
\texttt{s-inf} won't hurt the fairness (though it will change the order of 
answers). We swapped them back so that answers are produced in a more natural 
order.

@@
